---
title: "Getting Started with h5lite"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{h5lite}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(h5lite)

# We'll use a temporary file for this guide.
file <- tempfile(fileext = ".h5")
```

## Introduction

The `h5lite` package provides a simple, lightweight, and user-friendly interface for reading and writing HDF5 files. It is designed for R users who want to save and load R objects (vectors, matrices, arrays) to an HDF5 file without needing to understand the low-level details of the HDF5 C API.

This guide will walk you through a common use case: simulating experimental data, saving it to an HDF5 file along with metadata, and then reading it back for analysis.

### What is an HDF5 File?

Think of an HDF5 file as a self-contained file system. It's a single file on disk that can hold an organized hierarchy of your data. The three most important concepts are:

*   **Groups:** These are like folders or directories. You use them to organize your data. Groups can contain other groups to create a nested structure.
*   **Datasets:** These are like files in a file system. A dataset stores your actual data, such as a vector, matrix, or multi-dimensional array. Every dataset resides inside a group.
*   **Attributes:** These are small, named pieces of metadata that you can attach to either groups or datasets. They are perfect for storing extra information like units, descriptions, or configuration parameters.

`h5lite` is designed to make working with this structure feel natural to an R user.

## 1. Writing Datasets

The primary function for writing data is `h5_write()`. It creates a **dataset** inside the HDF5 file and automatically handles:

*   Creating the HDF5 file itself if it doesn't exist.
*   Creating parent **groups** as needed (like the command `mkdir -p`).
*   Overwriting any existing dataset at the same path.

Let's start by writing a matrix of simulated sensor readings to a dataset.

```{r write_matrix}
# A 3x4 matrix of sensor data
sensor_data <- matrix(rnorm(12, mean = 25, sd = 0.5), nrow = 3, ncol = 4)

h5_write(file, "experiment_1/sensor_readings", sensor_data)
```

That's it! You've just created an HDF5 file and stored a matrix in it.

> **Helpful Tip:** Notice the name `"experiment_1/sensor_readings"`. `h5lite` automatically created the group `experiment_1` before creating the dataset `sensor_readings` inside it.

### Specifying Data Types

By default, `h5lite` saves numeric data as `double` (64-bit float). If you need to control the on-disk storage type for efficiency, use the `dtype` argument. Let's save some integer identifiers.

```{r write_integer}
trial_ids <- 1L:12L
h5_write(file, "experiment_1/trial_ids", trial_ids, dtype = "integer")
```

### Writing Scalars

To write a single value (a scalar), you must specify `dims = NULL`.

```{r write_scalar}
h5_write(file, "experiment_1/run_id", "run-abc-123", dims = NULL)
```

## 2. Inspecting the File

Now that we've written some data, how do we see what's in the file?

### Listing Objects

`h5_ls()` lists the objects (groups and datasets) in the file. By default, it lists everything recursively.

```{r ls_recursive}
h5_ls(file)
```

To see only the top-level objects, use `recursive = FALSE`.

```{r ls_flat}
h5_ls(file, recursive = FALSE)
```

### Checking Dimensions and Types

You can inspect a dataset's properties without reading all of its data. This is useful for very large datasets.

*   `h5_dim()` returns the dimensions in R's standard, column-major order.
*   `h5_typeof()` returns the underlying HDF5 storage type.

```{r inspect_properties}
dim(sensor_data)
h5_dim(file, "experiment_1/sensor_readings")

h5_typeof(file, "experiment_1/trial_ids")
```

## 3. Reading Data

To read data back into R, use `h5_read()`. It automatically handles transposing the data from HDF5's row-major order to R's column-major order and restores the correct dimensions.

```{r read_data}
read_sensor_data <- h5_read(file, "experiment_1/sensor_readings")

print(read_sensor_data)

# Verify that the object is identical to the original
all.equal(sensor_data, read_sensor_data)
```

> **Safety First:** `h5lite` reads all numeric HDF5 types (integers, floats, etc.) into R's `numeric` (double-precision) vectors. This is an intentional design choice to prevent integer overflow, a common bug when reading data from other systems.

## 4. Working with Metadata (Attributes)

Attributes are small pieces of metadata attached to datasets or groups. They are perfect for storing things like units, configuration parameters, or version info.

Let's add some attributes to our `sensor_readings` dataset.

```{r write_attributes}
# Add a scalar string attribute for units
h5_write_attr(file, "experiment_1/sensor_readings", "units", "celsius", dims = NULL)

# Add a numeric vector attribute for calibration coefficients
h5_write_attr(file, "experiment_1/sensor_readings", "calibration", c(1.02, -0.5))
```

You can list and read attributes using `h5_ls_attr()` and `h5_read_attr()`.

```{r read_attributes}
h5_ls_attr(file, "experiment_1/sensor_readings")

units <- h5_read_attr(file, "experiment_1/sensor_readings", "units")
print(units)
```

## 5. Handling Special R Types

`h5lite` has special support for some of R's unique data types.

### Factors

When you write an R `factor`, `h5lite` automatically saves it as a native HDF5 `ENUM` type, preserving both the integer values and the character labels.

```{r handle_factors}
conditions <- as.factor(sample(c("control", "treatment_A", "treatment_B"), 12, replace = TRUE))

h5_write(file, "experiment_1/conditions", conditions)

# Let's check the on-disk type
h5_typeof(file, "experiment_1/conditions")

# Read it back - it's a perfect match!
read_conditions <- h5_read(file, "experiment_1/conditions")

identical(conditions, read_conditions)
```

### Raw Data

To store binary data, use an R `raw` vector and specify `dtype = "opaque"`. This stores the data as a sequence of bytes.

```{r handle_raw}
binary_blob <- as.raw(c(0xDE, 0xAD, 0xBE, 0xEF))
h5_write(file, "experiment_1/binary_config", binary_blob, dtype = "opaque")

read_blob <- h5_read(file, "experiment_1/binary_config")
identical(binary_blob, read_blob)
```

## 6. Managing File Contents

### Overwriting

`h5lite` follows an "overwrite-by-default" philosophy. If you write to an existing path, the old data is replaced.

```{r overwrite_data}
h5_read(file, "experiment_1/run_id")

# Overwrite with a new value
h5_write(file, "experiment_1/run_id", "run-xyz-987", dims = NULL)

h5_read(file, "experiment_1/run_id")
```

### Deleting Objects

You can explicitly delete datasets, groups, or attributes.

```{r delete_objects}
# Delete a single dataset
h5_delete(file, "experiment_1/trial_ids")

# Delete an attribute
h5_delete_attr(file, "experiment_1/sensor_readings", "calibration")

# Delete an entire group and all its contents
h5_delete_group(file, "experiment_1/binary_config") # This will fail, as it's a dataset
h5_create_group(file, "old_data/logs")
h5_delete_group(file, "old_data")

h5_ls(file, recursive = TRUE)
```

```{r cleanup}
# Clean up the temporary file
unlink(file)
```
